Interpretable Neural Networks using EAGGA

### 2025-03-10 paper/main.tex
more bullet on interpretability, mostly evaluation

(Why Interpretability is desirable?)
- in many ways desirable, as per \citet[pp. 3-4]{Zach2019InterpretabilityOD}, e.g.
  * \textbf{gaining trust}: most notable use case, e.g. medical diagnosis: predictions need to be validated by doctors \citep[p. 1]{review_NN_interpretability}
  * \textbf{model debugging}
    - model is optimised w.r.t. loss and judged on accuracy
    - might have unexpected performance drops in certain situations, makes model unreliable \citep[1B]{survey_NN_interpretability}
    - having an interpretable model + model induction process can help find ways to improve accuracy + reliability
  * \textbf{scientific understanding}: when only models can make sense of increasingly complex data anymore, interpretability enables extraction of learnt knowledge encoded
    in the model to make it accessible + reliable to humans (also mentioned here \citep[p. 1]{review_NN_interpretability})
  * \textbf{subconcious bias}, e.g. loan approval: must ensure that decision is not discriminatory
  * \textbf{regulatory}
    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability} or
    - drug approval processes in situations where a machine learning model discovered a new drug, process needs to be transparent for the regulator to approve \citep[1B]{survey_NN_interpretability}

@@

>>>>

(Why Interpretability is desirable?)
- in many ways desirable, as per \citet[pp. 3-4]{Zach2019InterpretabilityOD}, e.g.
  * \textbf{gaining trust}: most notable use case, e.g. medical diagnosis: predictions need to be validated by doctors \citep[p. 1]{review_NN_interpretability}
  * \textbf{model debugging}
    - model is optimised w.r.t. loss and judged on accuracy
    - might have unexpected performance drops in certain situations, makes model unreliable \citep[1B]{survey_NN_interpretability}
    - having an interpretable model + model induction process can help find ways to improve accuracy + reliability
  * \textbf{scientific understanding}: when only models can make sense of increasingly complex data anymore, interpretability enables extraction of learnt knowledge encoded
    in the model to make it accessible + reliable to humans (also mentioned here \citep[p. 1]{review_NN_interpretability})
  * \textbf{subconcious bias}, e.g. loan approval: must ensure that decision is not discriminatory
  * \textbf{regulatory}
    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability}, \citep[1B]{survey_NN_interpretability} or
    - drug approval processes in situations where a machine learning model discovered a new drug, process needs to be transparent for the regulator to approve \citep[1B]{survey_NN_interpretability}

@@

(Evaluation)
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated by based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies
